---
title: "Problem Set 3 (CSS)"
subtitle: "Due by 06/15/26"
execute:
  echo: true
  eval: false
editor: source
format: html
---


```{r}
library(needs)

needs(sf,
      tidyverse,
      readr,
      ggplot2,
      dplyr,
      osmdata,
      osm
      )
```

                            
**Download the official Leipzig shapefile (including the city districts) from the [Leipzig Open Data Portal](https://www.leipzig.de/fileadmin/mediendatenbank/leipzig-de/Stadt/02.1_Dez1_Allgemeine_Verwaltung/12_Statistik_und_Wahlen/Geodaten/Leipzig_Ortsteile_UTM33N.zip).**



```{r}
leipzig <- sf::st_read(dsn = "Data/ot.shp") # dsn = Data Set Name

head(leipzig)

ggplot(leipzig) +
  geom_sf(aes(fill = Name)) +
  #scale_fill_brewer(palette = "Set3") +  # or use another palette you like
  theme_minimal() +
  labs(fill = "Name",
       title = "Leipzig Map by Name")
```


1. Mark and plot the following three locations on the map: *Your favorite bar*, *your favorite restaurant* and *your house*  [^1] (*2 Points*)

[^1]: Feel free to make up locations if you're not comfortable sharing real ones or if you don't live in Leipzig. You can also use a different map area (e.g., Halle or Markkleeberg). This exercise is purely for practice purposes.



```{r}

home <- st_sf(
  name = "🏠",
  geometry = st_sfc(st_point(c(12.368751698679835, 51.34574546722337)),
                    crs = 4326)
)


pekar <- st_sf(
  name = "🍕",
  geometry = st_sfc(st_point(c(12.332844247771963,51.337003362550604)),
                    crs = 4326)
)

 
skala <- st_sf(
  name = "🍺",
  geometry = st_sfc(st_point(c(12.366968427661668,51.34147953528001)),
                    crs = 4326)
)

home <- st_transform(home, st_crs(leipzig))
pekar <- st_transform(pekar, st_crs(leipzig))
skala <- st_transform(skala, st_crs(leipzig))

# Combine into one sf object
locations <- rbind(skala, pekar, home)

ggplot() +
  geom_sf(data = leipzig, fill = "lavender", color = "lightgray") +
  geom_sf(data = locations, color = "purple", size = 3) +
  geom_sf_text(data = locations, aes(label = name), nudge_y = 0.0018, size = 3.5) +
  theme_minimal() +
  labs(title = "Your Leipzig POIs", subtitle = "🏠 Home, 🍕 Pekar, 🍺 Skala")

```



2. Calculate the straight-line distance (in kilometers) between your home and your favorite bar.(*1 Point*)



```{r}

# Calculate distance in meters
dist_m <- st_distance(home, skala)

# Convert to kilometers
dist_km <- as.numeric(dist_m) / 1000
```



3. Use R to verify whether the three points you chose are located within the same neighborhood boundaries. (*1 Point*)



```{r}
# Spatial join: assign each point its neighborhood from leipzig
home_quarter <- st_join(home, leipzig)
pekar_quarter <- st_join(pekar, leipzig)
skala_quarter <- st_join(skala, leipzig)

# Extract the neighborhood name (assuming the column is called 'Name')
home_name <- home_quarter$Name
pekar_name <- pekar_quarter$Name
skala_name <- skala_quarter$Name

# Check if all names are the same
 home_name == pekar_name && pekar_name == skala_name
 



```


4. Identify the centroid (geometric center) of your restaurants neighborhood. Is your favorite restaurant more than 800 meters away from it? (*2 Points*)


```{r}

restaurant_neighborhood <- leipzig[leipzig$Name == pekar_quarter$Name, ]

neighborhood_centroid <- st_centroid(restaurant_neighborhood)

distance_m <- st_distance(pekar, neighborhood_centroid)

as.numeric(distance_m) > 800



```






5. Download the [Migration statistics](https://opendata.leipzig.de/dataset/wanderungen-unrevidiert-jahreszahlen-kleinraumig) of Leipzigs neighborhoods from the Open Data Portal. Determine the ranking position of your home neighborhood in terms of *Innerstädtische Zuzüge* (i.e., people moving in from other parts of Leipzig). (*4 Points*)



```{r}

zuzuege_innerstaedtisch <- read_csv("data/Bevölkerungsbewegung_Wanderungen.csv") |>
  filter(Sachmerkmal == "Innerstädtische Zuzüge")|>
  pivot_longer(
    cols = starts_with("20"),  # matches all year columns like 2015, 2016, ..., 2024
    names_to = "Jahr",
    values_to = "Zuzuege"
  ) |>
  mutate(Jahr = as.integer(Jahr),
         Zuzuege = as.integer(Zuzuege))  # convert to numeric for ranking

leipzig_data <- left_join(x = leipzig,                  # Zusammenfügen der Daten über den Ortsteil
                          y = zuzuege_innerstaedtisch,
                          by = c("Name" = "Gebiet"))





# Filter for the most recent year
migration_2024 <- leipzig_data |>
  filter(Jahr == 2024) |>
  arrange(desc(Zuzuege)) |>
  mutate(Rank = row_number())

# Show your home quarter's rank
migration_2024|> 
  filter(Name == home_name)
```



6. Search the [OSM-Wiki](https://wiki.openstreetmap.org/) for how *Spätis* could be identified in Leipzig. Plot the identified Spätis on a map. (*5 Points*)




```{r}


# Ensure your Leipzig shapefile is in WGS84
leipzig_wgs <- st_transform(leipzig, 4326)

# Now get the bbox in lon/lat
bb_coords <- st_bbox(leipzig_wgs)

# 2. Query for shop=convenience
spatis <- opq(bbox = bb_coords) |>
  add_osm_feature(key = "shop", value = "convenience") |>
  osmdata_sf()

# Alternative: Sortieren nach Uhrzeit - oder halt beide

late_spatis <- spatis$osm_points |>
  filter(grepl("24/7|22:|23:|00:|01:", opening_hours, ignore.case = TRUE))

# 3. Optional: filter names that include "Späti", "Spätkauf"
spatis_named <- spatis$osm_points |>
  filter(grepl("Späti|Spätkauf|Kiosk", name, ignore.case = TRUE))

spatis_named_late <- spatis$osm_points |>
  filter(
    grepl("Späti|Spätkauf|Kiosk", name, ignore.case = TRUE) &
    grepl("24/7|22:|23:|00:|01:", opening_hours, ignore.case = TRUE)
  )


# View a sample of the opening hours
head(spatis_named$name, 10)

spatis_named <- st_transform(spatis_named, st_crs(leipzig))

intersects_matrix <- st_intersects(spatis_named, leipzig, sparse = FALSE)

intersects_any <- apply(intersects_matrix, 1, any)

# Subset points inside Leipzig polygons
spatis_named <- spatis_named[intersects_any, ]


ggplot() +
  geom_sf(data = leipzig, fill = NA, color = "gray60") +
  geom_sf(data = spatis_named, aes(geometry = geometry), color = "purple", size = 2) +
  labs(title = "Spätis in Leipzig",
       subtitle = "Identified by name filter from shop=convenience",
       caption = "Source: OpenStreetMap via osmdata") +
  theme_minimal()


```




7. Determine which city district has the highest density of Spätis. (*2 Points*)

```{r}
# Ensure both layers use the same CRS
spatis_named <- st_transform(spatis_named, st_crs(leipzig))

# Spatial join: assign each Späti to a district
spatis_districts <- st_join(spatis_named, leipzig, join = st_within)

spatis_per_district <- spatis_districts |>
  count(Name) |>
  rename(spati_count = n) |>
  arrange(desc(spati_count))


leipzig_counts <- leipzig |>
  left_join(st_drop_geometry(spatis_per_district), by = "Name")

leipzig_counts <- leipzig_counts |>
  mutate(
    area_km2 = st_area(geometry) / 10^6,
    density = spati_count / area_km2
  )




leipzig_counts <- leipzig_counts |>
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 10^6,  # convert to numeric square kilometers
    density = spati_count / area_km2
  )

ggplot(leipzig_counts) +
  geom_sf(aes(fill = density)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Density of Spätis by District in Leipzig",
       fill = "Spätis/km²") +
  theme_minimal()

```




8. Create a 1 kilometer buffer around your home. How many Spätis are located within that buffer? (*1 Point*)

```{r}
home_buffer <- st_buffer(home, dist = 1000)

spatis_named_proj <- st_transform(spatis_named, st_crs(home_buffer))

spatis_within_buffer <- spatis_named_proj[st_within(spatis_named_proj, home_buffer, sparse = FALSE), ]

n_spatis <- nrow(spatis_within_buffer)


```





9. What are the potential limitations of the upper data collection and operationalization method in measuring Späti accessibility? (*3 Points*)

10. Assume each person visits the nearest Späti to their home. Determine which Späti has the largest catchment area. (*3 Points*)

```{r}

spatis_proj <- st_transform(spatis_named, 32633)


bb_proj <- st_transform(leipzig, 32633)


voro_raw <- st_voronoi(st_union(spatis_proj), envelope = st_as_sfc(st_bbox(bb_proj)))


voro_polygons <- st_collection_extract(voro_raw, "POLYGON")


voro_sf <- voro_sf |> 
  mutate(spati_id = row_number()) |>        # ID for polygons
  bind_cols(spatis_named |>                  # bind Späti attributes
              st_drop_geometry() |> 
              select(name))     






# Calculate area in square kilometers
voro_clipped <- voro_clipped |>
  mutate(area_m2 = st_area(geometry)) |>
  mutate(area_km2 = set_units(area_m2, km^2)) |>
  mutate(area_km2 = as.numeric(area_km2))

largest_voro <- voro_clipped |> 
  filter(area_km2 == max(area_km2))

print(largest_voro)

ggplot() +
  geom_sf(data = leipzig, fill = NA, color = "grey40", linetype = "dashed") +
  geom_sf(data = voro_clipped, aes(fill = factor(spati_id)), color = "black", alpha = 0.6) +
  geom_sf(data = spatis_proj, color = "purple", size = 2) +
  scale_fill_viridis_d(option = "magma", guide = "none") +  # discrete color palette, no legend
  labs(
    title = "Voronoi Cells Around Spätis in Leipzig",
    caption = "Data: OpenStreetMap via osmdata"
  ) +
  theme_minimal()

```



